Modeling Lake Trophic State: A Data Mining Approach

### 2014-09-04 inst/doc/manuscript.Rmd
cleaned up uneeded files, update on RMD

---
title: "Modeling Lake Trophic State: A Data Mining Approach"
bibliography: LakeTrophicModelling.bib
layout: preprint
csl: ecology.csl
documentclass: elsarticle
output: 
  pdf_document:
    template: elsarticle.latex
    keep_tex: true
    fig_caption: true
author: 
  - name: Carl Boettiger
    affiliation: cstar
    email: cboettig(at)gmail.com
    footnote: Corresponding author
  - name: First Coauthor 
    affiliation: shortcode
address: 
  - code: cstar
    address: | 
      Center for Stock Assessment Research,
      110 Shaffer Rd,
      Santa Cruz, CA 95050, USA 
  - code: shortcode
    address: |
      Full address of First Coauthor
abstract: | 
          Write the abstract of the paper here
---
<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Lake Trophic Modelling}
%\usepackage[vmargin=0.5in,hmargin=0.5in]{geometry}
%\usepackage{times}
-->

##Introduction
Cyanobacteria are an important taxonomic group associated with harmful algal blooms in lakes. Understanding the drivers of cyanobacteria presence has important implications for lake management and for
 the protection of human and ecosystem health.  Chlorophyll a concentration, a measure of the biological productivity of a lake, is one such driver and is largely, although not exclusively, determined
 by nutrient inputs.   As nutrient inputs increase, productivity increases and lakes transition from low trophic state (e.g. oligotrophic) to higher trophic states (e.g. hypereutrophic).  These broad
trophic state classifications are associated with ecosystem health and ecosystem services/disservices (e.g. recreation, aesthetics, fisheries, and harmful algal blooms).  Thus, models of trophic state
 might be used to predict things like cyanobacteria.

@@

##Data and Modeling Methods

**Data**
We utilize four primary sources of data for this study. These are outlined below and in Table 1.

1. National Lakes Assessment (NLA) 2007: The NLA data were collected during the summer of 2007 and the final data were released in 2009.  With consistent methods and metrics collected at 1056 location
s across the conterminous United States (Map 1), the NLA provides a unique opportunity to examine broad scale patterns in lake productivity.  The NLA collected data on biophysical measures of lake wat
er quality and habitat.  For this analysis we primarily examined the water quality measurements from the NLA [@usepa2009national].
2. National Land Cover Dataset (NLCD) 2006: The NLCD is a nationally collected land use land cover dataset.  We collected total land use land cover and total percent impervious surface within a 3 kilo
meter buffer surrounding the lake to examine larger landscape-level effects [@homer2004development;@xian2009updating].
3. Modeled lake morphometry: Various measures of lake morphometry (i.e. depth, volume, fetch, etc.) are important in understanding lake productivity, yet many of these data are difficult to obtain for
 large numbers of lakes over broad regions.  To add this information we modeled lake morphometry [@hollister2010volume;@hollister_predicting_2011;@Lakmorpho2014;@NLMDInPrep].
4. Estimated Cyanobacteria Biovolumes: Cyanobacteria biovolumes is a truer measure of Cyanobacteria dominance than abundance as there is great variability in the size within and between species.  To a
ccount for this, Beaulieu *et al.* [-@beaulieu2013nutrients] used literature values to estimate biovolumes for the taxa in the NLA.  They shared this data and we have summed that information on a per-
lake basis.

##References

>>>>

---
title: 'Modeling Lake Trophic State: A Data Mining Approach'
author:
- affiliation: aed
  email: hollister.jeff(at)epa.gov
  footnote: Corresponding author
  name: Jeffrey W. Hollister
- affiliation: aed
  name: W. Bryan Milstead
- affiliation: aed
  name: Betty J. Kreakie
address:
- address: "US Environmental Protection Agency Office of Research and Development National
    Health and Environmental Effects Research Laboratory Atlantic Ecology Division 27
    Tarzwell Drive  Narragansett, RI, 02879, USA  "
  code: aed
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    template: elsarticle.latex
  html_document: null
  word_document: null
csl: plos.csl
documentclass: elsarticle
layout: review
linenumbers: no
bibliography: LakeTrophicModelling.bib
abstract:  Productivity of lentic ecosystems has been well studied and it is 
  widely accepted that as nutrient inputs increase,  productivity increases and
  lakes transition from low trophic state (e.g. oligotrophic) to higher trophic 
  states  (e.g. eutrophic). These broad trophic state classifications are good 
  predictors of ecosystem health and ecosystem services/disservices (e.g. recreation,
  aesthetics, fisheries, and harmful algal blooms). While the relationship between 
  nutrients and trophic state provides reliable predictions, it requires *in situ* 
  water quality data in order to paramterize the model.  This limits the application 
  of these models to lakes with existing and, more importanly, available water quality
  data.  To expand our ability to predict in lakes without water quality data, we take
  advantage of the availability of a large national lakes water quality database, 
  land use/land cover data, lake morphometry data, other universally available data, 
  and modern data mining approaches to build and assess models of lake tropic state 
  that may be more universally applied.  We use random forests and random forest 
  variable selection to identify variables to be used for predicting trophic state and 
  we compare the performance of two models of trophic state (as determined by 
  chlorophyll a concentration).   The first model estimates trophic state with *in situ* 
  as well as universally available data and the second model uses universally available 
  data only. For each of these models we used three separate trophic state categories, 
  for a total of six models. Overall accuracy for the *in situ* and universal data 
  models ranged from xx% to xx% and xx, xx, and xx described the most variation in 
  trophic state.  For the universal data only models, Overall accuraccy ranged from xx% 
  to xx% and xx, xx, and xx described the most variation in trophic state. Lastly, it 
  is believed that the presence and abundance of cyanobacteria is strongly associated 
  with trophic state.  To test this we examine the association between estimates of 
  cyanobacteria biovolume and the measured and predicted trophic state.  Expanding 
  these preliminary results to include cyanobacteria taxa indicates that cyanobacteria 
  are significantly more likely to be found in highly eutrophic lakes. These results 
  suggest that predictive models of lake trophic state may be improved with additional 
  information on the landscape surrounding lakes and that those models provide 
  additional information on the presence of potentially harmful cyanobacteria taxa. 
---
<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Modeling Lake Trophic State: A Data Mining Approach}
-->

#Introduction
Cyanobacteria are an important taxonomic group associated with harmful algal blooms in lakes. Understanding the drivers of cyanobacteria presence has important implications for lake management and for
 the protection of human and ecosystem health.  Chlorophyll a concentration, a measure of the biological productivity of a lake, is one such driver and is largely, although not exclusively, determined
 by nutrient inputs.   As nutrient inputs increase, productivity increases and lakes transition from low trophic state (e.g. oligotrophic) to higher trophic states (e.g. hypereutrophic).  These broad
trophic state classifications are associated with ecosystem health and ecosystem services/disservices (e.g. recreation, aesthetics, fisheries, and harmful algal blooms).  Thus, models of trophic state
 might be used to predict things like cyanobacteria.

@@

#Data and Modeling Methods

##Data

We utilize four primary sources of data for this study. These are outlined below and in Table 1.

1. National Lakes Assessment (NLA) 2007: The NLA data were collected during the summer of 2007 and the final data were released in 2009.  With consistent methods and metrics collected at 1056 location
s across the conterminous United States (Map 1), the NLA provides a unique opportunity to examine broad scale patterns in lake productivity.  The NLA collected data on biophysical measures of lake wat
er quality and habitat.  For this analysis we primarily examined the water quality measurements from the NLA [@usepa2009national].
2. National Land Cover Dataset (NLCD) 2006: The NLCD is a nationally collected land use land cover dataset.  We collected total land use land cover and total percent impervious surface within a 3 kilo
meter buffer surrounding the lake to examine larger landscape-level effects [@homer2004development;@xian2009updating].
3. Modeled lake morphometry: Various measures of lake morphometry (i.e. depth, volume, fetch, etc.) are important in understanding lake productivity, yet many of these data are difficult to obtain for
 large numbers of lakes over broad regions.  To add this information we modeled lake morphometry [@hollister2010volume;@hollister_predicting_2011;@Lakmorpho2014;@NLMDInPrep].
4. Estimated Cyanobacteria Biovolumes: Cyanobacteria biovolumes are a truer measure of Cyanobacteria dominance than abundance as there is great variability in the size within and between species.  To a
ccount for this, Beaulieu *et al.* [-@beaulieu2013nutrients] used literature values to estimate biovolumes for the taxa in the NLA.  They shared this data and we have summed that information on a per-
lake basis.

#References
